Detecting Motifs in Knowledge Graphs using Compression

### 2018-09-06 kgmotifs/kgmotifs.tex
deadline approaching

\usepackage{aaai19}    % Required
\usepackage{times}     % Required
\usepackage{helvet}    % Required
\usepackage{courier}   % Required
\usepackage{url}       % Required
\usepackage{graphicx}  % Required
\frenchspacing         % Required
\setlength{\pdfpagewidth}{8.5in}  % Required
\setlength{\pdfpageheight}{11in}  % Required

@@

\begin{abstract}
\noindent We introduce a method to detect \emph{network motifs} in knowledge graphs. Network motifs are useful patterns or meaningful subunits of the graph that recur frequently. We extend the common definition of a network motif to coincide with a \emph{basic graph pattern}, commonly used in the knowledge graph literature. We introduce an approach, inspired by recent work for simple graphs, for inducing these for a given knowledge knowledge graph, and show that the motifs returned reflect the basic structure of the graph. Specifically we show that in random graphs, no motifs are found, and that when we insert a motif artificially, it can be detected, even if as few as 10 instances are inserted. Finally, we show the results of motif induction on three real-world knowledge graphs, and show that our method provides good suggestions for motifs that reflect the structure of the graph in an intuitive way.
\end{abstract}

\noindent \emph{Knowledge graphs} are an extremely versatile way of storing knowledge. They allow knowledge to be encoded without a predefined schema, they allow different amounts of data to be encoded for each known entity and they allow different modalities to be integrated seamlessly \cite{wilcke2017knowledge}. This versatility comes at a price. For a given knowledge graph, it can be difficult to see the forest for the trees: how is the graph structured at the lowest level? What kind of things can I ask of what types of entities? What are small, recurring patterns that might represent a novel insight into the data?

In the domain of unlabeled simple graphs, \emph{network motifs} \cite{milo2002network} were introduced as a tool to provide insight into local graph structure. Network motifs are small subgraphs whose frequency in the graph is unexpected with respect to a \emph{null model}. That is, for a given graph dataset $G$ and small graph $M$, we count the frequency $f(M, G)$: how often $M$ occurs in $G$ as a subgraph. We define a probability distribution over graphs $p^\text{null}(\G)$, and estimate the probability that a graph sampled from $p^\text{null}$ contains more instances of $M$ than observed in our data: $p^\text{null}(f(M, \G) \geq f(M, G))$, where $\G$ is a random variable representing a graph. If this probability is very low (commonly, below 0.05), we consider $m$ a motif. \footnotemark

@@

Unfortunately, estimating this probability usually requires repeating the subgraph count on many samples from the null model. To avoid this costly operation, \cite{bloem2017large} introduces  an alternative method, using \emph{compression} as a heuristic for motif relevance: the better a motif compresses the data, the more likely it is to be meaningful. It is shown that this principle can be implemented in a similar sort of hypothesis test for a null model, allowing a comparable workflow to classical motif analysis.

\footnotetext{This procedure has the structure of a hypothesis test, but it is important not to interpret it as \emph{statistical evidence} for the meaningfulness of the motif. The only thing it shows (in a frequentist statistical sense) is that $p^\text{null}$ is not the true source of the data. This is usually not a surprise: we are rarely able to model all aspects of a realistic data-generating process in a single distribution. The $p$-values used in motif analysis should be interpreted strictly as \emph{heuristics}. See also \cite{bloem2017large}.}

@@

We perform several experiments to show that our method returns meaningful subgraphs. First we test the intuition that a random graph shoudl contain no motifs. We also show that when we artifically insert motifs into a random graph, we can then detect these as motifs, sometimes with as few as 10 instances inserted into the graph. Finally, we show the results of motif analysis on three real-world knowledge graphs.

@@

When we need to encode a single choice from a finite set $S$ of options, we will often use the code with length $\log |S|$, corresponding to a uniform distribution on $S$. In some cases, we can allow codes with multiple codewords for a single object. If a code $L$ has multiple codewords for some object  $x$, we may indicate the choice for a particular codeword by a parameter $a$ as $L(x; a)$.

\paragraph{Relevance testing} We will use the MDL principle to perform a hypothesis test. Assume we have some data $x$ and a null hypothesis that it was sampled from distribution $p^\text{null}$. Let $L^\text{null}$ be the corresponding codelength. A simple but crucial result, known as the \emph{no-hypercompression inequality} \cite[p103]{grunwald2007minimum} tells us that the probability of sampling data $x$ from $p^\text{null}$ that can be described shorter than $L^\text{null}(x)$ by $k$ or more bits, \emph{using any code} is less than $2^{-k}$.
Thus, a standard MDL hypothesis test would consist of stating the null hypothesis in terms of a code $L^\text{null}$, designing a code (before seeing the data) which compresses the data better than $L^\text{null}$ by, say, 10 bits and rejecting the null hypothesis with confidence $2^{-10}$. For a longer, more intuitive explanation of this principle in pattern induction, we refer the reader to \cite[Section~2]{bloem2017large}.

We must note again, that when we use this procedure to find motifs, we are not providing evidence for the hypothesis that the motif is ``correct.'' The only thing we are statistically proving is that the null-model is incorrect. Nevertheless, hypothesis testing has shown to be a valuable heuristic in motif analysis.

\paragraph{Common codes}
In the construction of our codes for graphs, we require some simpler codes as building blocks. First, when we store any positive integer $n$, we do so with the code corresponding to the distribution $p^\N(n) = 1/(n(n+1))$, and denote it $L^\N(n)$. For nonnegative numbers we add 1 to the argument. for the full range of integers, we add an extra bit for the sign, and then use the first code for negative integers and the second for positive ones.

@@

Let $f(A, B)$ be the frequency of symbol $A$ in sequence $B$.We then store the complete sequence using the code corresponding to the following distribution:
\begin{align*}
&p^\text{PY}_{\alpha,d}(S) = \prod_{i \in [1,k]} \text{PY}_{\alpha,d}(S_i\mid S_{1:i-1}) \\
&\text{PY}_{\alpha,d}(S_i \mid S') = 
\begin{cases}
\frac{\alpha - d |m(S)|}{|m(S)| + \alpha} & \text{if} f(S_i, S') = 0 \\ 
\frac{f(S_i, S') - d}{|m(S)| + \alpha} & \text{otherwise}
\end{cases}
\end{align*}
See \cite{de2016names} for a more intuitive explanation. In all experiments we use $\alpha = 0.5$, $d=0.1$. We will refer to the total for a sequence  codelength under this procedure as $L^PY(S)$

\section{Method}

We will start with some formal definitions of our basic ingredients. To comfortably define probabilities and codes on graphs, we will slightly deviate from conventional definitions. Specifically, we will analyse the \emph{structure} of knowledge graphs only, ignoring any meaning they have outside the graph. For instance, we will distinguish between links with different labels, but we will ignore any internal structure of the label itself (such as shared IRI prefixes). Under this assumption we can assume that the node and link labels of our knowledge graphs are simply natural numbers.\footnotemark This leads to the following definition.

@@

@@

We will first assume that a target pattern $M$ is given for the data $G$ and that we have a set of instances $\I$. Moreover, we require that all instances in $\I$ are mutually disjoint: no two subgraphs defined by member of $\I$ may share an edge, but nodes may be shared. In most settings we do not actually have a given $M$ and $\I$ and we will need to search for them. We describe a simple and search algorithm in Section~\ref{section:search}.

\label{section:relevance-test}

\section{Null model}

\label{section:null-model}

The most common null model for classical motif analysis is the degree-sequence model (also known a the configuration model \cite{newman2010networks}): a uniform distribution over all graphs with a particular degree sequence. We extend this to knowledge graphs by also including the degree of each relation: that is, the frequency with which it occurs in the tripleset. Let a \emph{degree sequence} $D$ of length $n$ be a triple of three integer sequences: $(D^\text{in}, D^\text{rel}, D^\text{out})$. If $D$ is the degree sequence of a graph, then node $i$ has $D^\text{in}_i$ incoming links,  $D^\text{out}_i$ outgoing links and for each relation $r$, there are $D^\text{rel}_r$ triples.

@@

 \paragraph{Encoding D} In order to encode a graph with $L^\text{EL}_D$, we must first encode $D$. \footnote{Or, equivalently, to make $p^\text{EL}$ a complete distribution on all graphs, we must provide it with a prior on $D$.} For each of the three sequences in $D$ we use the following model:
\begin{align*}
 p(D) &= \prod_i q^{\N}(D_i) \\
 L(D) &= - \sum_i \log q^{\N}(D_i)
\end{align*}

@@

@@

We note a few aspects of this code. Firstly, it achieves its compression by reducing the number of links it has to store. For each additional instance, it pays the cost of storing the instance, and any variable nodes and links, but it saves $|E_M|$ links in storing the template. The precise trade-off depends on the choice of base code, and on the structure of the rest of the graph (some edges are cheaper to store than others, for some codes).

Nevertheless, we can make a few broad observations: patterns with many variable nodes and links are more costly to store than fully labeled patterns. On the other hand, a pattern with many variable nodes and links is more likely to have many instances, so that \emph{if} it results in a small amount of compression per instance, the large number of instances may cause it to compress a lot. For the variable nodes and links, we use a DM model for each specific node and link. This means that the more structure there is to the label given to a particular part of the pattern, the better we compress. For instance, suppose a particular node of the pattern is always labeled with either a node corresponding to YES or NO, in equal proportion, we will require only a bit per instance to record those labels. If, however, all nodes in the graph are equally likely to appear in the position of the variable, we will pay approximately $\log |V_G|$ bits per instance. If only a single node appears in the position, we pay only a a constant number of bits to record the entire sequence of instances. Thus, to our code, non-variable nodes are just a highly regular type of variable node. The distinction is only meaningful in the search algorithm.

@@

\subsection{Encoding motif instances}

To encode a list of instances $\I$ of a given pattern $M$, we generalize  the idea of the edgelist model described above. To understand the following ,it may be instructive to think of a single triple as an instance for the pattern
\[
\texttt{?n1 ?rel ?n2}.
\]
 We can now think of the edgelist model as storing a sequence of instances for this pattern, which together make up the entire knowledge graph.

 To generalize this notion to arbitrary patterns, to be defined for a given template graph, we define the \emph{degree constraint} $D^\I$ of a list of instamces for a given pattern as follows: for each variable node $i$ in the pattern, the degree sequence provides an integer sequence $D^i$ of length $v_G$, indicating how often each node in the completed knowledge graph takes that position in the pattern. Similarly, for each variable link $j$ in the pattern, the degree sequence provides an integer sequence $C^j$ of length $r_G$ indicating for each relation how often it takes that position in the pattern.

@@

Note that if we store a graph with the pattern $\texttt{?n1 ?rel ?n2.}$, we obtain an empty template graph, and the motif code achieves the same codelength as the edgelist model, up to a small constant amount for storing the pattern.

For a given graph and pattern, we can simply find the complete list of instances using a graph patter search. Since we require a slightly different semantics than standard graph pattern matchers, we adapt the DualIso algorithm \cite{saltz2014dualiso} for knowledge graph matching. Before computing the motif code, we prune the list of instances provided by this search iterating over the instances and removing and instance that produces a triple also produced by an earlier instance.

Since graph matching encompasses subgraph isomorphism, it is known to be NP-Complete. To guard against rare patterns that produce long-running searches we terminate all searches after 5 seconds, returning  only those matches that were found within the time limit.

@@

@@

Ultimately, we want to find any motifs that have a high log-factor for a given graph $G$. Since we can readily compute the log-factor, any black-box optimization algorithm can be used to search the space of all possible motifs. For the sake of simplicity, we will use basic simulated annealing here. The starting pattern is always a single random triple from the graph, with its relation made a variable. We define seven possible transition from one pattern to another:
\begin{description}
	\item[Extend (0.1)] Choose an instance of the pattern and an adjacent triple not part of the instance. Add the triple to the pattern.
	\item[Make a node a variable (3)] Choose a random constant node, and turn it in to a variable node.
	\item[Make an edge a variable (3)] Choose a random constant edge label, and turn it in to a variable (always introducing a new variable).
	\item[Make a variable node constant (2)] Choose a random variable node and turn it into a constant. Take the value from a random instance.
	\item[Make a variable edge constant (2)] Choose a random variable edge and turn it into a constant. Take the value from a random instance.
	\item[Remove edge (3)] Remove a random edge from the pattern, ensuring that it stays connected.
	\item[Couple (1)] Take two distinct edge variables, which for at least one instance hold the same value and turn them into a single variable.
\end{description}

The values in brackets are the relative weights of each transition type (chosen by trial and error to keep the algorithm stable). We choose a transition at random and compute a new pattern. If the transition cannot be made (for instance, there are no constant node to make variable) or if the resultant pattern is in some way invalid, we sample a new transition.

Once a new pattern has been sampled, we compare its codelength under the motif model to that of the previous sample. If the codelength is lower, we continue with the new pattern. If the codelentgh is longer, we continue with the new sample with probability $\alpha$ or return to the previous pattern otherwise. We use $\alpha = 0.5$ in all experiments.

We store all patterns seen and their scores. In order to exploit all available processor cores, we run several searches in parallel. After each the searches are finished, we tajke the top 1000 patterns from each and sort them by motif codelength. If two searches return different codelengths\footnotemark for a motif we take the lowest value.

\footnotemark{This is possible if the pattern search is cut off due to the 5 second time limit.}

\section{Experiments}

To validate the method, we first test it on random graphs. We sample a directed graph with a given number of nodes $n$ and edges $m$, with no self-connections and multiple edges (that is, we sample from from the $G(n, m)$ Erd\H{o}s-Renyi model). We then label the nodes uniformly at random with one of the relations in $0, \ldots, r$. To make the dimensions realistic we base them on thos of the Mutag dataset used later: $n=23644$, $m74567$, $r=24$.

We then take one randomly chosen pattern, and insert $k$ instances of the pattern into the graph. We run the experiment for $k=0$, $k=10$ and $k=100$.

\section{Acknowledgments}

Withheld for purposes of blind reviewing.
% This publication was supported by the Amsterdam Academic Alliance Data Science (AAA-DS) Program Award to the UvA and VU Universities.

>>>>

\usepackage{aaai19}    % Required
\usepackage{times}     % Required
\usepackage{helvet}    % Required
\usepackage{courier}   % Required
\usepackage{url}       % Required
\usepackage{graphicx}  % Required
\frenchspacing         % Required
\setlength{\pdfpagewidth}{8.5in}  % Required
\setlength{\pdfpageheight}{11in}  % Required
\usepackage{makecell}
\usepackage{arydshln}
\usepackage{float}

@@

\begin{abstract}
\noindent We introduce a method to detect \emph{network motifs} in knowledge graphs. Network motifs are useful patterns or meaningful subunits of the graph that recur frequently. We extend the common definition of a network motif to coincide with a \emph{basic graph pattern}, commonly used in the knowledge graph literature. We introduce an approach, inspired by recent work for simple graphs, for inducing these for a given knowledge knowledge graph, and show that the motifs returned, reflect the basic structure of the graph. Specifically we show that in random graphs, no motifs are found, and that when we insert a motif artificially, it can be detected, even if as few as 10 instances are inserted. Finally, we show the results of motif induction on three real-world knowledge graphs, and show that our method provides good suggestions for motifs that reflect the structure of the graph in an intuitive way.
\end{abstract}

\noindent \emph{Knowledge graphs} are an extremely versatile way of storing knowledge. They allow knowledge to be encoded without a predefined schema, they allow different amounts of data to be encoded for different entities and they allow different modalities to be integrated seamlessly \cite{wilcke2017knowledge}. This versatility comes at a price. For a given knowledge graph, it can be difficult to see the forest for the trees: how is the graph structured at the lowest level? What kind of things can I ask of what types of entities? What are small, recurring patterns that might represent a novel insight into the data? Answering these questions could benefit problem domains like graph simplification, graph navigation and schema induction.

In the domain of unlabeled simple graphs, \emph{network motifs} \cite{milo2002network} were introduced as a tool to provide insight into local graph structure. Network motifs are small subgraphs whose frequency in the graph is unexpected with respect to a \emph{null model}.

% That is, for a given graph dataset $G$ and small graph $M$, we count the frequency $f(M, G)$: how often $M$ occurs in $G$ as a subgraph. We define a probability distribution over graphs $p^\text{null}(\G)$, and estimate the probability that a graph sampled from $p^\text{null}$ contains more instances of $M$ than observed in our data: $p^\text{null}(f(M, \G) \geq f(M, G))$, where $\G$ is a random variable representing a graph. If this probability is very low (commonly, below 0.05), we consider $m$ a motif. \footnotemark

@@

Unfortunately, estimating this probability usually requires repeating the subgraph count on many samples from the null model. To avoid this costly operation, \cite{bloem2017large} introduces  an alternative method, using \emph{compression} as a heuristic for motif relevance: the better a motif compresses the data, the more likely it is to be meaningful.

%It is shown that this principle can be implemented in a similar sort of hypothesis test for a null model, allowing a comparable workflow to traditional motif analysis.

%\footnotetext{This procedure has the structure of a hypothesis test, but it is important not to interpret it as \emph{statistical evidence} for the meaningfulness of the motif. The only thing it shows (in a frequentist statistical sense) is that $p^\text{null}$ is not the true source of the data. This is usually not a surprise: we are rarely able to model all aspects of a realistic data-generating process in a single distribution. The $p$-values used in motif analysis should be interpreted strictly as \emph{heuristics}. See also \cite{bloem2017large}.}

@@

We perform several experiments to show that our method returns meaningful subgraphs. First we test the intuition that a random graph should contain no motifs. We also show that when we artificially insert motifs into a random graph, we can then detect these as motifs, sometimes with as few as 10 instances inserted into the graph. Finally, we show the results of motif analysis on three real-world knowledge graphs.

@@

% When we need to encode a single choice from a finite set $S$ of options, we will often use the code with length $\log |S|$, corresponding to a uniform distribution on $S$. In some cases, we can allow codes with multiple codewords for a single object. If a code $L$ has multiple codewords for some object  $x$, we may indicate the choice for a particular codeword by a parameter $a$ as $L(x; a)$.

\paragraph{Relevance testing} We will use the MDL principle to perform a hypothesis test. Assume we have some data $x$ and a null hypothesis that it was sampled from distribution $p^\text{null}$. Let $L^\text{null}$ be the corresponding codelength. A simple but crucial result, known as the \emph{no-hypercompression inequality} \cite[p103]{grunwald2007minimum} tells us that the probability of sampling data $x$ from $p^\text{null}$ that can be described shorter than $L^\text{null}(x)$ by $k$ or more bits, \emph{using any code} is less than $2^{-k}$.
Thus, an MDL hypothesis test would consist of stating the null hypothesis in terms of a code $L^\text{null}$, designing a code (before seeing the data) which compresses the data better than $L^\text{null}$ by, say, 10 bits and rejecting the null hypothesis with confidence $2^{-10}$. For a longer, more intuitive explanation of this principle in pattern induction, we refer the reader to \cite[Section~2]{bloem2017large}.

Note that when we use this procedure to find motifs, we are not providing evidence for the hypothesis that the motif is ``correct'' \cite[Section~2]{bloem2017large}.
\paragraph{Common codes}
In the construction of our codes for graphs, we require some simpler codes as building blocks. First, when we store any positive integer $n$, we do so with the code corresponding to the distribution $p^\N(n) = 1/(n(n+1))$, and denote it $L^\N(n)$. For nonnegative numbers we add 1 to the argument. for the full range of integers, we add an extra bit for the sign, and then use the first code for negative integers and the second for positive ones.

@@

Let $f(A, B)$ be the frequency of symbol $A$ in sequence $B$.We then store the complete sequence using the code corresponding to the following distribution:
\begin{align*}
&p^\text{PY}_{\alpha,d}(S) = \prod_{i \in [1,k]} \text{PY}_{\alpha,d}(S_i\mid S_{1:i-1}) \\
&\text{PY}_{\alpha,d}(S_i \mid S') = 
\begin{cases}
\frac{\alpha - d |m(S)|}{|m(S)| + \alpha} & \text{if} f(S_i, S') = 0 \\ 
\frac{f(S_i, S') - d}{|m(S)| + \alpha} & \text{otherwise}
\end{cases}
\end{align*}
See \cite{de2016names} for a more intuitive explanation. In all experiments we use $\alpha = 0.5$, $d=0.1$. We will refer to the total codelength for a sequence under this procedure as $L^{PY}(S)$

\section{Method}
To comfortably define probabilities and codes on graphs, we will slightly deviate from conventional definitions. Specifically, we will analyse the \emph{structure} of knowledge graphs only, ignoring any meaning they have outside the graph. Under this assumption we can assume that the node and link labels of our knowledge graphs are simply natural numbers.\footnotemark~This leads to the following definition.

@@

\begin{table}[H]x
\centering
\tiny
\begin{tabular}{r r p{0.5\linewidth}}
\hline
log-factor & frequency & \\
\hline
\multicolumn{3}{c}{Dogfood , top 5 by log-factor ($>100$ positive)}\\
\hline	
 83119.1 & 9344 & \makecell{\texttt{?n1 dc:creator ?n2.} \\\texttt{?n2 rdfs:label ?n3.} \\\texttt{?n2 foaf:name ?n3.} } \\ 
 \hdashline
 81075.1 & 10322 & \makecell{\texttt{?n1 foaf:maker ?n2.} \\\texttt{?n2 rdfs:label ?n3.} \\\texttt{?n2 foaf:name ?n3.} } \\
  \hdashline 
 77692.6 & 5234 & \makecell{\texttt{?n2 rdf:type ?n3.} \\\texttt{?n2 rdfs:label ?n1.} \\\texttt{?n2 foaf:name ?n1.} } \\ 
  \hdashline
 75632.3 & 3536 & \makecell{\texttt{?n1 swc:isPartOf ?n2.} \\\texttt{?n2 swc:hasPart ?n1.} } \\ 
  \hdashline
 75592.4 & 3536 & \makecell{\texttt{?n1 swc:isPartOf ?n2.} \\\texttt{?n2 ?p3 ?n1.} } \\ 
\hline
\multicolumn{3}{c}{Dogfood, top 3 by frequency}\\
\hline	
 -3032.5 & 137008 & \makecell{\texttt{?n1 rdf:\_1 ?n2.} \\\texttt{?n1 rdf:\_2 ?n4.} \\\texttt{?n1 rdf:\_3 ?n3.} } \\ 
  \hdashline
 -3251.4 & 109315 & \makecell{\texttt{?n1 swc:isPartOf ?n3.} \\\texttt{?n2 swc:isPartOf ?n3.} } \\ 
 \hdashline
 2089.0 & 92165 & \makecell{\texttt{?n2 swc:isPartOf ?n3.} \\\texttt{?n3 swc:hasPart ?n1.} } \\ 
\hline
\multicolumn{3}{c}{AIFB, top 5 by log-factor ($>100$ positive)}\\
\hline	
 79234.0 & 7728 & \makecell{\texttt{?n1 ?p3 ?n2.} \\\texttt{?n2 ?p4 ?n1.} } \\ \hdashline
 61310.4 & 4154 & \makecell{\texttt{?n1 swrs:publication ?n2.} \\\texttt{?n2 ?p3 ?n1.} } \\ \hdashline
 57641.1 & 3965 & \makecell{\texttt{?n1 swrs:publication ?n2.} \\\texttt{?n2 swrs:author ?n1.} } \\ \hdashline
 57603.1 & 3965 & \makecell{\texttt{?n1 swrs:author ?n2.} \\\texttt{?n2 ?p3 ?n1.} } \\ \hdashline
 42110.1 & 17512 & \makecell{\texttt{?n1 swrs:publication ?n3.} \\\texttt{?n3 swrs:author ?n2.} \\\texttt{?n3 swrs:author ?n1.} \\\texttt{?n2 swrs:publication ?n3.} } \\ 
\hline
\multicolumn{3}{c}{AIFB, top 3 by frequency}\\
\hline
 2287.7 & 198668 & \makecell{\texttt{?n1 swrs:publication ?n3.} \\\texttt{?n1 swrs:publication ?n2.} } \\ \hdashline
 7822.1 & 191083 & \makecell{\texttt{?n1 swrs:author ?n2.} \\\texttt{?n2 swrs:publication ?n3.} } \\ \hdashline
 1264.0 & 190636 & \makecell{\texttt{?n1 swrs:isAbout ?n3.} \\\texttt{?n2 swrs:isAbout ?n3.} } \\ 
\hline
\multicolumn{3}{c}{Mutag, top 5 by log-factor ($87$ positive)}\\
\hline
 178304.4 & 18634 & \makecell{\texttt{?n1 mtg:\_hasAtom ?n3.} \\\texttt{?n1 mtg:\_hasBond ?n2.} \\\texttt{?n2 mtg:\_inBond ?n3.} } \\ \hdashline
 97237.8 & 9189 & \makecell{\texttt{?n1 mtg:\_hasAtom ?n2.} \\\texttt{?n2 mtg:\_charge ?n3.} } \\ \hdashline
 90447.5 & 18634 & \makecell{\texttt{?n1 mtg:\_hasBond ?n2.} \\\texttt{?n2 mtg:\_inBond ?n4.} \\\texttt{?n2 mtg:\_inBond ?n3.} } \\ \hdashline
 79052.0 & 9317 & \makecell{\texttt{?n1 mtg:\_hasBond ?n2.} \\\texttt{?n2 rdf:type ?n3.} } \\ \hdashline
 58973.9 & 7892 & \makecell{\texttt{?n1 mtg:\_hasAtom ?n2.} \\\texttt{?n2 rdf:type ?n3.} } \\ 
\hline
\multicolumn{3}{c}{Mutag, top 3 by frequency}\\
\hline
 1025.7 & 341143 & \makecell{\texttt{?n2 mtg:\_hasAtom ?n3.} \\\texttt{?n2 mtg:\_hasBond ?n1.} } \\ \hdashline
 -374.5 & 178837 & \makecell{\texttt{?n1 mtg:\_charge ?n3.} \\\texttt{?n2 mtg:\_charge ?n3.} } \\ \hdashline
 -2088.0 & 52985 & \makecell{\texttt{?n1 rdf:type owl:Class.} \\\texttt{?n3 rdfs:subClassOf ?n2.} \\\texttt{?n4 ?p5 owl:Class.} \\\texttt{?n4 rdfs:subClassOf ?n2.} } \\ 
\hline
\end{tabular}
\caption{Results of the experiment on real-world data. For each experiment we also report the number of motifs found with a positive log-factor.}
\label{table:topk}
\end{table}

@@

We will first assume that a target pattern $M$ is given for the data $G$ and that we have a set of instances $\I$. Moreover, we require that all instances in $\I$ are mutually disjoint: no two subgraphs defined by member of $\I$ may share an edge, but nodes may be shared.

\paragraph{Null model}
The most common null model for classical motif analysis is the degree-sequence model (also known a the configuration model \cite{newman2010networks}): a uniform distribution over all graphs with a particular degree sequence. We extend this to knowledge graphs by also including the degree of each relation: that is, the frequency with which it occurs in the tripleset. Let a \emph{degree sequence} $D$ of length $n$ be a triple of three integer sequences: $(D^\text{in}, D^\text{rel}, D^\text{out})$. If $D$ is the degree sequence of a graph, then node $i$ has $D^\text{in}_i$ incoming links,  $D^\text{out}_i$ outgoing links and for each relation $r$, there are $D^\text{rel}_r$ triples.

@@

 \paragraph{Encoding D} In order to encode a graph with $L^\text{EL}_D$, we must first encode $D$. \footnote{Or, equivalently, to make $p^\text{EL}$ a complete distribution on all graphs, we must provide it with a prior on $D$.} For each of the three sequences in $D$ we use the following model:
\[
 p(D) = \prod_i q^{\N}(D_i)\;\;\;\;\;
 L(D) = - \sum_i \log q^{\N}(D_i)
\]

@@

\begin{table}[tbp]
\small
\begin{tabular}{r r p{4cm} r}
\hline
log-factor & frequency & &  \\
\hline
220360.2 & 12138 & \makecell{\texttt{?n1 foaf:maker ?n2.} \\\texttt{?n2 foaf:made ?n1.}} & D  \\ \hdashline
 3157.0 & 1011 & \makecell{\texttt{?n1 ?p2 "false".}}  & M \\ \hdashline
  3150.2 & 985 & \makecell{\texttt{?n1 ?p2 "true".}} & M \\ \hdashline
   12871.8 & 8308 & \makecell{\texttt{?n1 rdf:type ?n2.} \\\texttt{?n1 swrs:year ?n3.} \\\texttt{?n4 swrs:publication ?n1.} }& A \\ 
\hdashline
\end{tabular}
\caption{Selected motifs.The last column indicates the dataset.}
\label{table:selected}
\end{table}

@@

@@

\paragraph{Encoding motif instances} To encode a list of instances $\I$ of a given pattern $M$, we generalize  the idea of the edgelist model described above. To understand the following ,it may be instructive to think of a single triple as an instance for the pattern \texttt{?n1 ?rel ?n2}. We can think of the edgelist model as storing a sequence of instances for this pattern, which together make up the entire knowledge graph.

 To generalize this notion to arbitrary patterns, to be defined for a given template graph, we define the \emph{degree constraint} $D^\I$ of a list of instances for a given pattern as follows: for each variable node $i$ in the pattern, the degree constraint provides an integer sequence $D^i$ of length $v_G$, indicating how often each node in the completed knowledge graph takes that position in the pattern. Similarly, for each variable link $j$ in the pattern, the degree sequence provides an integer sequence $C^j$ of length $r_G$ indicating for each relation how often it takes that position in the pattern.

@@

Note that if we store a graph with the pattern \texttt{?n1 ?rel ?n2} we obtain an empty template graph, and the motif code achieves the same codelength as the edgelist model, up to a small constant amount for storing the pattern.

For a given graph and pattern, we can simply find the complete list of instances using a graph pattern search. Since we require a slightly different semantics than standard graph pattern matchers, we adapt the DualIso algorithm \cite{saltz2014dualiso} for knowledge graph matching. Before computing the motif code, we prune the list of instances provided by this search iterating over the instances and removing and instance that produces a triple also produced by an earlier instance. To guard against rare patterns that produce long-running searches we terminate all searches after 5 seconds, returning  only those matches that were found within the time limit.

@@

\begin{figure}[tb]
  \centering
    \includegraphics[width=\linewidth]{synthetic-plot.pdf}
    \caption{The result of the random graph experiment. We sort the motifs by their score in the $k=10$ experiment and plot their frequency and log-factor.}
    \label{figure:synthetic}
\end{figure}

@@

Ultimately, we want to find any motifs that have a high log-factor for a given graph $G$. Since we can readily compute the log-factor, any black-box optimization algorithm can be used to search the space of all possible motifs. For the sake of simplicity, we will use basic simulated annealing here. The starting pattern is always a single random triple from the graph, with its relation made a variable. We define seven possible transition from one pattern to another: \\
\textbf{Extend (2)} Choose an instance of the pattern and an adjacent triple not part of the instance. Add the triple to the pattern.\\
\textbf{Make a node a variable (3)} Choose a random constant node, and turn it in to a variable node.\\
\textbf{Make an edge a variable (3)} Choose a random constant edge label, and turn it in to a variable (always introducing a new variable).\\
\textbf{Make a variable node constant (2)} Choose a random variable node and turn it into a constant. Take the value from a random instance.\\
\textbf{Make a variable edge constant (2)} Choose a random variable edge and turn it into a constant. Take the value from a random instance.\\
\textbf{Remove edge (3)} Remove a random edge from the pattern, ensuring that it stays connected.\\
\textbf{Couple (1)} Take two distinct edge variables, which for at least one instance hold the same value and turn them into a single variable.

The values in brackets are the relative weights of each transition type (chosen by trial and error). We choose a transition at random using the normalized weights as probabilities and compute a new pattern. If the transition cannot be made (for instance, there are no constant node to make variable) or if the resultant pattern is in some way invalid, we sample a new transition.

Once a new pattern has been sampled, we compare its codelength under the motif model to that of the previous sample. If the codelength is lower, we continue with the new pattern. If the codelength is longer, we continue with the new sample with probability $\alpha$ or return to the previous pattern otherwise. We use $\alpha = 0.5$ in all experiments.

We store all patterns seen and their scores. In order to exploit all available processor cores, we run several searches in parallel. After each the searches are finished, we take the top 1000 patterns from each and sort them by motif codelength.

 \begin{figure}[tb]
  \centering
    \includegraphics[width=\linewidth]{synthrep-plot.pdf}
    \caption{The result of the repeated random graph experiment. Color and size show the number of matches of the pattern after pruning. Plot titles show the graph dimensions before adding instances.}
    \label{figure:synthrep}
\end{figure}
\section{Experiments}

\paragraph{Random graphs} To validate the method, we first test it on random graphs. We sample a directed graph with a given number of nodes $n$ and edges $m$, with no self-connections and multiple edges (that is, we sample from from the $G(n, m)$ Erd\H{o}s-Renyi model). We then label the nodes uniformly at random with one of the relations in $0, \ldots, r$. To make the dimensions realistic we base them on those of the Mutag dataset used later: $n=23644$, $m74567$, $r=24$.

We then take one randomly chosen pattern, and insert $k$ instances of the pattern into the graph. We run the experiment for $k=0$, $k=75$ and $k=150$. The results are shown in Figure~\ref{figure:synthetic}. For $k=0$, as expected, we find no patterns with positive compression. For $k=75$, we find ...such patterns and for $k=150$ we find ... . While we only inserted one pattern we can expect to find more than one, since its subpatterns may also become motifs. However, the pattern we inserted should receive the highest log-factor.

Of course, this experiment only tests a single pattern. To see the effect of multiple random patterns, we repeat the experiment many times, sampling both the pattern and the random graph. 
To sample the pattern we first sample a random number of nodes $n$ from $U(3, 6)$, the uniform distribution over the integer range $(3, 6)$ (including both end points). We then sample a random number of links $m$ from $U(n, n^2 - n)$, and sample a random directed graph from $G(n, m)$. We make $U(0, n)$ nodes and $U(0, m)$ links into variables, choosing constants for the rest uniformly from the data. If the pattern is disconnected, we reject and sample again.

We sample a random graph as in the previous experiments, using the dimensions from the three real world datasets used later. We then add $k$ instances of the motif to the graph and compute the log-factor of the sampled pattern (we do not conduct a search).

We let $k$ range from 0 to 200, and repeat the experiment 25 times for each $k$, sampling a new graph and pattern each time. The results are shown in Figure~\ref{figure:synthrep}. We observe first that under this ad-hoc sampling regime, we produce some patterns that create only very few instances in the graph, after overlapping instances are pruned. Since it is no surprise that these don't allow significant compression, we plot these as small points so that they don't obscure the other points.

We see that most of the other instances---those that generate enough non-overlapping instances---can be retrieved as motifs. As the number of added instances increases, the proportion of motifs with separate instance that cannot be retrieved diminishes.

\paragraph{Real-world data} Finally, we will test our method on real-world dat, to confirm that the motifs found coincide with our intuition. We test three datasets: The semantic Web dogfood dataset \cite{moller2007recipes} ($n=7611$,$m=242256$,$r=170$) describing researchers and publications in the Semantic Web domain, the AIFB dataset \cite{bloehdorn2007kernel} ($n=8285$,$m=29226$,$r=47$) describing the structure of the AIFB institute, and the Mutag RDF dataset \footnotemark ($n=23644$,$m=74567$,$r=24$), describing a set of carcinogenic and non-carcinogenic molecules both in structure and properties.

\footnotetext{Distributed as an example dataset with the DL-Learner framework. \cite{lehmann2009dl}}

For all datasets we run 32 parallel searches, with 3125 iterations per search. For each dataset we report the top 5 motifs by log-factor, and the top 3 motifs by frequency. Note that these are the top 10 motifs for frequency in the set of 32000 high-log-factor motifs resulting from the parallel search. The supplemental materials show the top 100 motifs for both criteria.

The patterns with negative score but high frequency show the pitfalls our code avoids. For instance, in the top motif by frequency for the Dogfood dataset, we see a four-node pattern with two edges fanning in to a single node. These are caused by a single node having the same ingoing relation to many other nodes. Clearly such a pattern is very frequent in the data, but capturing just three of these incoming links does not constitute a satisfying motif, and indeed such patterns receive a negative log-factor.

Much of what the motif code picks up on is redundancy in the original data. For instance, in the AIFB data both the \texttt{swrs:publication} relation and its inverse \texttt{swrs:author} are always included. Extracting these into a motif is simple way of achieving compression. In fact, the AIFB data caontains so many of these relation pairs that the two-node loop with variable labels is the highest scoring motif. In the Dogfood data, we see similar patterns emerge.
